---
#########################################
# options for knitting a single chapter #
#########################################
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
    citation_package: biblatex
  bookdown::html_document2: default
  bookdown::word_document2: default
documentclass: book
bibliography: [bibliography/references.bib]
#chapter level bibliography
#bibliography: [bibliography/synth-ref.bib]
params:
  width: 70%
---

# Organisms associated with anthropogenic marine debris reported in the scientific literature globally from 1990 to 2019
\chaptermark{OAAMD Database}

```{r setup, include = F, }
library(tidyverse)
library(janitor)
library(purrr)

file_location <- '~/PhD/thesis_git/mac-thesis/data/biosec/'
```


```{r load files, include = F, }

pull_dfs <- c("organisms", "papers", "spatial", "taxonomy")
file_type <- c(".csv")

list_dfs <- pull_dfs %>%
  map(~ read.csv(paste0(file_location, ., file_type), encoding = "UTF-8")) %>%
  map(~ na_if(., "")) %>%
  map(~ remove_empty(., which = c("rows", "cols"))) %>%
  map(~ janitor::clean_names(.)) %>%
  map(~ rename_with(., ~str_remove(., "x_u_feff_")))
  
names(list_dfs) <- pull_dfs

list2env(list_dfs, envir = globalenv())

organisms_clean <- organisms %>%
  select(-rank) %>%
  mutate(scientific_name = fct_recode(as.factor(scientific_name),
                                      Actiniaria = "Actinaria",
                                      Amoebophrya = "Amoebophyra"))
#names need further cleaning

df_orgs <- left_join(organisms_clean, taxonomy)

```

```{r manual table creation, include = F}



```

## Abstract
	- How main measurement value is important (e.g. chlorophyll).
	- Why it is the choice measurement for that purpose.
	- How many records are included of main measurement and the spatiotemporal scope of their collection.
	- Primary methods used to collect and process data in source material.
	- Brief description of what source was used to find records.
	- Where to access the database.
	- What the database can be combined with and for which common uses.

### Corresponding author

```{r data summary table, echo = F}





```
- Tables
	- Summary table containing
		- design type(s): what type of database is it and why was it assembled
		- measurement type(s): what is the main measurement (eg. chlorophyll)
		- technology type(s): what technology was used to collect the data (e.g. spectroscopy)
		- factor type(s): how main measurement is primarily grouped
		- sample characteristic(s): other measurements in the dataset

- Background & Summary
	- Relevance of data. Why data exists. Compare to alternatives to metrics stored in data. 
	- Caveats to using the data. Types of data used alongside for analyses/ context.
	- General trends in data and their biological origin.
	- Example of these trends. 
	- Importance of outliers and other patterns in data to the users/collectors of it.
	- Example of program/project that relies on the data.
	- Alternative, less important origins for trends in the data. 
	- Describe what data is present.
	- Describe where the data came from: organisations, projects, people.
	- Figure explaining data collection process.
	- Suggestion for pre-processing needed prior to using the data for various applications. 
	- Anywhere else the data is offered.
	- Describe any ongoing updates or maintenance to that dataset. 
	- Explain link between static set and that set.

- Methods
	- Overview of methods used to collect the field data.
	- Explain where to find examples of each general method or a review paper describing various methods in detail. 
	- Overview of most common methods to process the field data. 
	- Brief comparison of these methods.
	- Mention of other methods possible to use but not included in the dataset and description of reason why they were excluded. 
	- Describe how methods many be affected by different common scenarios that present in the field and lab.
	- Delineate sources of data compiled into the database.
	- Describe how data were accessed.
	- Describe how data was filtered for relevancy.
	- Describe how data was made to fit into a standardised data resource.
	- Describe how ongoing maintenance or updating would be performed.

- Figures
	- Map of records in database
	- Record availability over time in a bar graph with a broad categorizing variable used for color
	- Diagram of process from data collection to final database.

- Data Records
	- Define the lowest level of record and what it describes (include def of identifier)
	- Describe how lowest records are linked through to other levels.
	- Describe highest level of record and why it collects the lower levels.
	- How metadata is linked to the dataset.
	- Any caveats to the above, like why multiple names of a species might be attached to a record.
	- Quality assurances for each record and how they were validated.

- Technical Validation
	- How data was cleaned and prepared in a way that needs to be acknowledged in any analysis.
	- How data quality issues were addressed in the records that were lower quality but still met the minimum standards to be included.
	- Any conversions of preparations that may be confusion to a user at first.
	- Null or missing value signifiers.

- Usage Notes
	- Where the data has been made available for use.
	- Suggest complementary datasets.
	- Suggest how to link the data through to those complementary sets.
	- Examples of how this type of data has been used in projects or publications in combination with other sources.
	- Common reasons that the dataset admin would be contacted.
	- How to contact the dataset admin.
	- Where to go for more information about the dataset.

- References
	- bibliography

- Data Citation
	- citation of dataset in stable repository

- Acknoledgements
	- Broad acknowledgment of data providers and collectors.
	- How to cite using info from this dataset (e.g. if single, cite source. if multi, cite dataset).
	- Encourage engagement with source material.
	- Where to find specific acknowledgements of sources.
	- Specific requirements in using specific subsets of the data (e.g. JTMD).

- Author Contributions
	- Which author performed which function in the creation of the dataset and paper.
	- Where to reference to get more specific description of contributions.

- Additional Information
	- Any tables, figures, or support material only available online or from a certain source.

- Competing Interests
	- None.

- Appendices
	- Table of detailed description of each authors contribution.





