---
#########################################
# options for knitting a single chapter #
#########################################

output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
    citation_package: biblatex
    extra_dependencies: ["booktabs"]
  bookdown::html_document2: default
  bookdown::word_document2: default
documentclass: book
params:
  width: 70%

#maintenance check code
bibliography: [bibliography/references.bib] 

#chapter only
#bibliography: [bibliography/biosec-ref.bib] 

---

```{r setup, include = F}

library(knitr)
library(tidyverse)
library(kableExtra)

```

```{r localFunctions, include = F}

lineBreaks <- function(.text = matDefine, .max_str_length = 30) {
  
  objType <- typeof(.text)
  
  if (objType == "list") {
    all_out <- "dummy"
  } else if (objType == "character") {
    
    all_out <- list()
    
    for (i in .text) {
      str_length <- stringr::str_length(i)
      if (str_length <= .max_str_length) {
        str_out <- i
      } else {
        split_start <- seq(from = 1, to = (str_length - 6), by = .max_str_length)
        split_end <- c((start_split[-1] - 1), str_length)
        split_strs <- stringr::str_sub(i, split_start, split_end)
        need_brks <- split_strs[-length(split_strs)]
        end_str <- split_strs[length(split_strs)]
        added_brks <- c(paste0(need_brks, "//-/n"), end_str)
        str_out <- paste(added_brks, collapse = "")
        all_out <- "dummy"
          }
    }
  } else {
    print("Input must be a character vector or dataframe with at least one character column.")
  }
  
  return(all_out)
  
}

```

# Fishing gear and other anthropogenic debris harbour invasive species

\chaptermark{Biosecurity concerns of AMD}


## Abstract  

Records of non-indigenous and invasive species in published scientific literature were found in all marine regions, except the Southern Ocean, and more often than would be expected in a comparable marine community. Comparisons were enabled by the organisms associated with AMD database of scientific literature published between 1992 and 2019 combined with taxonomic and invasion history data available on the World Register of Marine Species. Across the nine animal phyla recorded in the OAAMD, the proportion of formally designated invasive species in communities on anthropogenic marine debris (AMD) was over 60 times higher than in the overall marine community. Invasive species were found outside of their specialist-delineated native marine regions in over 85% of records. Of these, 80 records represented novel findings in marine regions outside of their known range. Records of invasive species were also significantly more common within 150km of an urban coastal area. Although the type of AMD was often not reported at the level of an individual organism by the studies synthesised, available data did suggest that NIS may more often be found on structurally complex debris items like fishing gear and synthetic fabrics than on flat or smooth plastics, metal, and natural materials.  


## Introduction  


Driftwood, kelp mats, and other types of naturally occurring debris floating at the ocean’s surface support diverse communities of organisms adapted to the specific habitat conditions they provide [@thielRaftingBenthicMacrofauna2003]. The long history of association between these communities and natural floating debris has moulded geographic distributions of these organisms along the paths of surface currents driven by prevailing winds [@luizSeafarersCastawaysEcological2015; @pfallerHitchhikingHighSeas2019]. Rafting, the process in which species are transported by floating debris, has also been shown to occur aboard the mass of anthropogenic marine debris (AMD) now floating alongside this natural mixture [@barnesRemoteIslandsReveal2005; @gilOceanicBarnaclesAct2016; @mirallesAlertCallingPort2018].

Since AMD presents a large and increasing amount [@koelmansAllNotLost2017; @lebretonEvidenceThatGreat2018] of solid surface area in the top layer of the ocean, it may also present a large and increasing amount of habitat capable of transporting associated organisms. This implication of increasing plastic production and pollution was hypothesised early on [@winstonDriftPlasticExpanding1982], as observational studies continued to find beach debris fouled frequently by species noted to be similar to those found on **Sargassum sp.** [@gregoryHazardsPersistentMarine1991] and on the beaches of extremely remote islands [@ryanAccumulationStrandedPlastic1988]. These early papers proposed that pristine coastlines, previously buffered from human impacts by sheer distance, would become vulnerable to anthropogenic disturbance and the introduction of harmful species both marine and terrestrial. 

The next thirty years of research show that the scope of impacts from rafting aboard AMD is potentially far more complex and pervasive. Fouling organisms, multicellular plants and animals which grow attached to the surface of an object immersed in water, have been found on AMD in every marine region [@citekey] on beaches both remote [@citekey] and highly urbanised [@citekey]. In oceans and seas, this fouling community has been shown to attract and support diverse communities of mobile predators and grazers [@citekey]. Among these associated organisms are those with histories of causing harm when established outside of their native range [@citekey]. Even at the microbial level, AMD has been shown to carry pathogens harmful to humans [@citekey], fish [@citekey], coral [@citekey], and other organisms [@citekey]. The distances travelled by AMD and these problematic species are much further than any recorded aboard natural substrate [@citekey].

### Natural versus artificial

This capacity for long-distance dispersal of hitch-hikers is a distinct feature of AMD. Natural materials and their abundance at the ocean's surface are often heavily influenced by the seasons [@citekey] and can be limited in how far they travel by physical and biological degradation [@citekey]. AMD can be added to the ocean year-round [@citekey]. Products manufactured from plastics, wood, rubber, and other artificial or enhanced materials are typically far more resistant to physical degradation [@citekey] and have few organisms known to be capable of breaking them down [@citekey]. This means that AMD can arrive and persist at times and in places where natural debris is highly unlikely. 

Direct comparisons of the organisms living on AMD with those on natural debris has also mostly disproved initial observations of similarity between the two. Cosmopolitan species like gooseneck barnacles (**Lepas sp.**) and certain cheliostome bryozoans like **Electra tenella** are common to both. The rest of the community, however, has often been shown to vary significantly [@citekey; @citekey]. Studies have shown that some invasive species and pathogenic microbes are significantly more common on AMD [@citekey], while others were found to raft only on AMD [@citekey].

### Dispersal potential

All organisms with the capacity to use AMD as habitat may be more able to shift the boundaries of their known ranges and disperse over very long distances. Abundant substrates, particularly around urban areas [@citekey] or after natural disasters [@citekey], could offer these organisms more chances and at more optimal times of year to establish populations in new, previously inaccessible, areas. This could help maintain more biodiversity as suitable habitats shift around with climate change. AMD associated organisms may be more likely to randomly find their new habitable range than those not able to utilize it. 

However, this same process may also increase the likelihood that invasive species spread to areas beyond the site of their initial incursion (secondary invasion) or to regions far beyond the reach of natural sources of population control. When the Tohuku-Oki earthquake caused a tsunami that flooded the area surrounding Fukushima Daiichi, Japan in 2011, entire communities of marine life were transported across the North Pacific on thousands of items ranging from household litter to entire piers [@Carlton2017]. No natural substrate, like driftwood, has even been documented making that same trip [@citekey]. These organisms, separated by tens of thousands of kilometers and hundreds of thousands of years of evolution, were only able to make the trans-Pacific journey because of AMD. Dozens of them were identified as species with histories of invasion [@citekey].

Numerous other studies have documented shorter distance travel aboard debris to new locations just outside of a species known range [@citekey; @citekey]. Through short hops from ports areas, seen in @Miralles2014, AMD has the potential to expand the spatial extent of incursions in these types of invasion hotspots into a large, low-density halo around the initial site. Likely due to its inherently random, omnipresent, and indirect character compared to other ways that humans transport organisms in the marine environment, AMD has not yet been soley implicated in any introduction of an invasive species. However, its documented ability to transport marine organisms merits its consideration among other modes of transporting potentially harmful organisms [@Audrézet2021], known as **introduction vectors**. 

### AMD as a vector

In the context of marine biosecurity, introduction vector refers to a way that humans transport species around oceans and seas. **Pathways** are the common trajectories taken by those vectors, such as high-traffic shipping lanes. Some species transported by vectors along pathways are designated by regulatory or management bodies as **invasive species**. This designation is generally given once it has been demonstrated that the species causes significant harm to human and/or natural systems. Invasive species are typically associated with a specific vector, or set of vectors, known to facilitate their dispersal. AMD could be considered a vector for each of the hundreds of species known to use it as habitat [@citekey].

A 2021 meta-analysis estimated that harms related to aquatic invasive species, combining both freshwater and marine environments, cost the global economy a total of 23 million USD in 2020 alone [@Cuthbert2021]. These costs may arise directly from mitigation efforts and lost revenues in fisheries [@Schwoerer2019] or damage to infrastructure [@Park1999], among other economic impacts. Beyond monetary costs, invasive species can change how ecosystems function, altering the composition of native species [@Jackson2017] and reducing the services those ecosystems provide to human communities [@Katsanevakis2014]. 

Characterisation of distinct introduction vectors facilitates a border-focussed biosecurity strategy, suggested to be the most efficient means of preventing and responding to incursions of NIS [@Hewitt2007]. Doing so for AMD requires delineating common pathways, identifying sinks along coastline where there is a higher concentrations of arrivals, describing the physical environment encountered by organisms transported aboard the vector and along the pathway, and compiling species known to use the vector [@Faulkner2020]. In the case of AMD as an introduction vector, this characterisation can be pulled from decades of tangential research in oceanography, materials science, and marine biology [@Audrézet2021; @Bergmann2015]. 

### Assessing the potential of AMD

To facilitate assessment of AMD as a vector, this study combined a 2021 database of published records of organisms associated with anthropogenic marine debris (OAAMD) [@McMains2022] with the World Register of Introduced Marine Species (WRiMS) [@Rius2020], to compare the frequency of invasive species found on AMD with that of the rest of the comparable marine environment. This combination of sources also enabled the assessment of spatial and methodological patterns in records of invasive species on AMD. Combined with a comprehensive list of invasive species found on AMD, this investigation aimed to delineate which species use this vector and characterise patterns in its pathways. 

### Material types

Not all substrates categorised as AMD provide the same type of habitat and certain substrates are more associated with certain pathways. These distinctions often do not fall along the same lines as those useful when sorting debris found on the beach. As such, the way AMD is categorised for the purposes of this study is different than in many others. The categorisation described below also accomodates reporting gaps in the literature and groups substrates into categories broad enough to be statistically useful.

#### Fishing gear

Nylon fishing nets, expanded polystyrene foam buoys, and synthetic rope were grouped alongside other diverse materials originating from fishing and aquaculture activities. Though they are made from a variety of polymers, with different chemical additives, and offer diverse surface properties to fouling organisms, this category shares two definitive characteristics. First, almost all of these materials will be discharged into the marine environment with an existing marine biofilm. This primes their surfaces for easy attachment of juvenile macro-organisms and increases the likelihood that they already have adult fouling organisms attached. Second, is that they are designed to be used outdoors in sunny, wet, and salty conditions. Previous studies have shown that these characteristics can or do manifest as distinct biological communities [@citekey; @citekey].

#### Megastructure

Large objects comprised of multiple named or unnamed materials were grouped as megastructures. These objects are likely to provide several different types of habitat, enabling niche partitioning for greater diversity. As a whole object, their tendency to resist degradation for long periods of time may enable these diverse communities to travel longer distances and develop into more complex community structures [@citekey]. There may also be enough reproductive adults to sustain stable populations over long journeys [@citekey]. This makes them distinct from smaller objects which may not have sufficient space to do so.

#### Plastic

Single material objects of indeterminant origin were often simply labelled plastic in their source paper. Where spetroscopy was used to accurately identify the polymer, this was rarely reported in association with the organism itself. This methodological quirk is mitigated by the fact that polymer type has been shown to have far less effect on the associated community as plastics weather and degrade [@citekey]. Several studies have shown that buoyancy, age, and season of discovery have a stronger effect on the plastic associated community than polymer type [@citekey]. The first two factors are even more rarely noted in the literature than polymer type. Differentiating between hard and foam plastic may be useful in future studies (soft plastics are rarely found fouled outside of the benthos), but there was not enough data to do so here.

#### Glass and ceramic

Glass bottles and ceramic fouling plates, which comprise the majority of this group, have both been shown to follow similar successional patterns [@citekey] due to their low surface energy. They present a relatively difficult surface for establishment when first introduced to a marine environment and are resistant to the surface pitting that facilitates settlement by macro-organisms. However, they both are fragile to impact and are prone to sinking if damaged. They are also far less likely than plastics to adsorb and absorb chemicals that may harm the associated community.

#### Fabrics

There was no reporting of fabric type in any of the studies in the OAAMD. However, common natural fibres are less water resistant [@citekey] and can therefore be assumed to be relatively rare at the oceans surface. Polyester and other artificial fibres, including heavily modified natural fibres like modal, were therefore combined into one group along with the potential inclusion of natural fibres which had not yet become too waterlogged to float. Despite the fibre, all fabric substrates can be assumed to provide ample gaps for the attachment of organisms and little resistance to folding or turning in ways expose attached organisms to a variety of environments and stressors.

#### Metal

Metals were never reported to their alloy or specific metallic content and were comparatively rare. Data availability determined this category. Different metals and their characteristic chemical interaction with the marine environment and organisms would suggest that separating them might improve understanding of material specific patterns. However, it can be reasonably assumed that metals provide a distinct habitat to that provided by the other categories. Including them in the analysis was seen to provide a useful basis for further targeted exploration because, despite their rarity in the OAAMD, they are often found in beach clean-ups [@citekey].

#### Other

This group contains records of rubber, paint chips, and objects described to their purpose where they could not be placed in any other category. Their dissimilarity means this tactic offers little information value each sub-type. Instead, it can be seen as a background catch-all for artificial substrates, suggesting an uncertain average for these materials.

#### Unknown

In some cases there was no available information on the object that the organism was found on. In these cases, where the study focussed entirely on AMD, it could only be assumed that the object was artificial or manufactured. This was kept separate from the "Other" category because almost all of the records in this group came from the Japanese Tsunami Marine Debris database. The source was considered to separate these two categories into distinct populations.

#### Natural

Kelp, driftwood, and seagrasses were sometimes mentioned in studies otherwise focussed on AMD. Each of these materials is known to have a different associated biological community, but there were not enough records to separate them out individually and still maintain comparable statistical inference to what was available for the other categories. Like the other category, this grouping is most useful to provide an uncertain average across these materials and for comparison to AMD. 


## Methods

This study combines information from four databases compiled fully cited peer-reviewed or specialist-verified records:

 - *OAAMD*: Organisms associated with anthropogenic marine debris [@McMains2022] synthesised over 20 years of published research in life- and materials-science to provide a collated database of recorded associations of organisms with floating and immersed anthropogenic marine debris and their locations; 
 - *WRiMS*: The World Register of Introduced Marine Species [@Rius2020] combined contemporary lists of species known to biosecurity agencies as invasive, managed, or potentially concerning from 64 countries (Supplementary data); 
 - *WoRMS*: The World Register of Marine Species [@WoRMSEditorialBoard2020] provided standardised nomenclature for each organism and was utilised as the most comprehensive and robust available reference of marine species known to science;
 - *Marineregions.org* [@citekey]: Compiled geospatial databases of marine regions from multiple gazettes into one source with stable resource identifiers accessible from a public API such that each spatial coordinate pair in the OAAMD could be associated with a Marine Ecoregion of the World (MEOW) [@citekey].

### Data compilation

The OAAMD provided the lowest level records for which the rest of the information was queried and added. WRiMS was used to indicate whether each recorded organism had a history of invasion. WoRMS provided the full taxonomic classification for each record according to general scientific consensus as of January 2021. Its functionality enabled older binomial names and classifications to be updated for the analysis and summary tables in this chapter. Marine ecoregions for each spatial coordinate pair in the OAAMD were pulled from the Marineregions.org API in May 2021. The same API was also used to convert countries with invasive species lists in WRiMS to Marine Ecoregions.

Lists of invasive species in WRiMS were labelled by the country or managed marine region that had designated the species as invasive, managed, or of concern. These three broad categories were lumped into one label "NIS" for this analysis because there were very few records of organisms in the OAAMD which were designated using the latter two categories. NIS was chosen to represent the three to accommodate the fact that not all of the species were fully listed as invasive.

Known contemporary geographic distributions for each species found in the OAAMD and listed an NIS in WRiMS were pulled from specialist-validated sources in April 2021. These sources, in order of preference were:
 - the Smithsonian’s NEMESIS database [Fofonoff2021];
 - WoRMS [WoRMSEditorialBoard2020];
 - CABI Invasive Species Compendium [@CABI2021];
 - and the Global Invasive Species Database [@GISD2021].  
 
These sources were supplemented, in some cases, by published scientific articles or theses (Supplemental data). Where active scientific debate or uncertainty about an organism’s native range was ongoing in 2020, the organism was listed as having no native range and excluded from analyses focused on regional patterns. These distributions were described using Marine Ecoregions of the World [@citekey]. 

### Material type

For the purposes of this analysis, material type was summarised into several broad categories based on the more varied assortment reported in the OAAMD. The categories, described earlier in this chapter, are again summarised in Table \@ref(tab:materialDefine). They represent an attempt to best group within data availability and factors known to influence fouling patterns.

```{r materialDefine, echo = F, out.width = params$width}

matTypes <- c("Plastic", "Fishing gear", "Megastructure", "Fabric", "Glass or ceramic", "Metal", "Other", "Unknown")

matDefine <- c("All petroleum-derived synthetic materials were not used in fishing gear.",
               "A subset of plastics that originated as the result of fishing or aquaculture.",
               "Complex large structures composed of multiple potential materials.",
               "Synthetic and artificial fabrics as often vaguely described in sources.",
               "Smooth and rigid non-petroleum derived artificial substrates.",
               "All materials identified as metal, or a type of.",
               "Broad and diverse category of materials each with a small number of records.",
               "Used where material type was unreported in the source.")

matData <- data.frame(matTypes, matDefine)

knitr::kable(matData,
      col.names = c("Material", "Definition"),
      caption = "Material type designations used to classify the broader range of materials found in the OAAMD to provide adequate sample sizes in the following statistical analyses.",
      caption.short = "Material type definitions",
      booktabs = T) %>%
  kable_styling(latex_options = "scale_down")

```

### Analysis

Statistical analysis and visualisations for this paper were completed in the R Statistical Computing environment (R Core Team, 2018) and tested the following hypotheses: 

 - H1: The percentage of NIS within the community recorded on AMD is not significantly higher than in the background marine community;
 - H2: The percentage of NIS within the community recorded on AMD is not significantly higher on debris within 150km of coastal urban area than outside of this radius;
 - H3: NIS are not found within the community on AMD more frequently in marine regions directly adjacent to or within the native range for the species.

### H1: Prevalence of invasive species in the recorded AMD community

The community recorded on AMD is determined by physical aspects of the environment, the biology of each organism, interactions between organisms, and biases inherent to scientific effort and publication. These may combine to impact how many species on a debris item are designated as invasive species, or NIS for the purpose of this analysis. The aim of this test was to determine whether the unique characteristics of AMD increase the prevalence of NIS above that found in the background marine community.

Randomly simulated communities of animals from nine phyla were created from the two separate lists of species: OAAMD and WoRMS. The percent of this community comprised of NIS was calculated and used to generate two separate frequency distributions. These were then compared using a chi-square test. This analysis did not account for regional factors, unavailable for the species not listed as NIS, and instead investigated overall global trends. 

Random, unweighted sampling without replacement was performed using the random number generator included as a basic function in the R programming language [@citekey]. The number of samples in each run was selected randomly from a normal distribution with the same mean, minimum, and maximum as species richness within studies in the OAAMD (minimum 6, mean = 42, max = 89). 10,000 communities were randomly generated in this way from each list of species. The percent of the total community which was identified as an NIS was then calculated.

This process was repeated 10,000 times for each of the two datasets, resulting in 20,000 calculations of the percent of NIS in simulated community. An ANOVA based on a logit-link linear model [@citekey] was used to test whether the distribution of percent NIS was significantly different between communities simulated from the OAAMD and WoRMS datasets. A logit model was selected because its assumptions align with the derived probabilities calculated during the simulations [@citekey].

### H1a: Increased likelihood of recording listed invasive species 

This process was performed again with the addition of a weighting value applied to the selection of species listed as NIS. The weighting value increased the likelihood that an NIS would be selected at random, relative to non-NIS species, from the list of species found in WoRMS. Weighting values were set at each of the numbers between 5 and 100 to generate a logit-link linear model of the relationship between likelihood of selecting an NIS and the resulting proportion of the simulated community that NIS represent. These values were selected to incorporate the high variability in percent NIS seen in the unweighted model and anchored around the overall ratio of NIS within the AMD and background marine communities compared (60:1, from memory, need to check). A statistically similar proportion of NIS to that found on artificial debris was inferred to be found where confidence intervals (2.5% and 97.5%) overlapped the mean proportion calculated from the 10,000 simulations on the AMD community.

### H2: Urban centres influence on prevalence of listed invasive species


Geographic coordinates for each record, provided within the OAAMD, allowed for the identification of records found within a 150km radius of the centre point of a named settlement with a population over 50,000 [@citekey]. Records within this radius were considered urban, according to the definition of large towns and urban centres provided by the Australian Bureau of Statistics [@citekey] A chi-squared analysis [@citekey] was used to test whether the number of records of listed invasive species was significantly higher than non-listed species on AMD within this radius.

### H3: Connectivity between the native range of NIS on debris and where it was found 


The scientifically established ranges of each NIS were compared to the regions where the organisms were recorded in the OAAMD database [@citekey]. This allowed records of NIS found within their native ranges to be excluded from the connectivity analysis. The “distance” between each remaining record and its nearest native or introduced range was then calculated as the fewest number of regions it would have “passed through” to travel between the two using a breath first algorithm in the iGraph implementation in R [@citekey]. Each record was assumed to have come from the nearest native region for that taxa. When distance ties occurred, both regions were assumed to have been the source. Secondary introductions, in which a species is transferred from its introduced range into another new region, and more complicated journeys across multiple regions were not included in the assumptions of this model. Though likely scenarios, the dataset used includes numerous sources of variability and error that would not support such a complex model.

Connection weight was quantified as the number of NIS records in region “A” likely originating from region “B”. Regions were grouped using the algorithm developed by Brandes et. al which maximises the strength of the connections within the group and minimises connections to nodes outside of the group [@citekey]. This analysis and the creation of visualisations were performed in the R implementation of iGraph [@citekey] using tidygraph [@citekey] and ggraph [@citekey]. Records from Japanese tsunami marine debris were excluded from this portion to achieve more balanced sample sizes.

## Results


Of the 6,542 organisms in the complete database of OAAMD, 792 belonged to species listed as an invasive, managed, or concerning species in the World Register of Invasive Marine Species (Rius et al., 2020) (Table XX, Supplementary material). Throughout the recorded time-period (1990-2019), the ratio of unique NIS to unique taxa stayed steady at around 5% (Figure 1). Records were concentrated in the NE Pacific Ocean (n = 684), due to the thorough study of debris related to the Japanese tsunami (Carlton et al., 2017). Outside of this study, the next most frequent regions were European territorial waters in the NE Atlantic Ocean (n = 33) and Mediterranean Sea (n = 28). The listed invasive species most frequently found are globally wide-spread and known to have broad environmental tolerances: Mytilus galloprovincialis [@citekey] (n = 291), Megabalanus rosa [@citekey] (n = 115), and Magallana gigas [@citekey] (n = 74). Listed taxa were most common on fishing gear (n =270) and megastructures (n = 200), the latter primarily associated with Japanese tsunami debris.

Plastics, the most frequently recorded material harbouring associated organisms (n = 2,506), showed a comparatively low presence (3%) of listed NIS within its associated community (Figure 2). By contrast, items categorised as fabrics (n = 856), showed a very high percentage of NIS within its community. Out of all records related to fabric, 13% were listed NIS, which was a close third behind fishing gear (20%) and megastructures (16%). Few records of NIS were found on hard surfaces like metal (n = 32) or glass and ceramic (n = 25).

Material type was missing for a high number of records in the OAAMD database overall (n = 773) and almost never recorded at the level of an individual organism, which precluded testing hypotheses about the material preferences of listed taxa. Location, however, was recorded for all the studies reporting the biological community on floating artificial substrates. While often not recorded at the level of an individual taxa (43.7% of records), most studies recorded location data at the study level on a discrete and delineated area of less than 200 km2. 

A general scientific consensus, within the sources listed in the methods, was also available for the native and introduced ranges of all but 6 out of 80 total listed species: 2 listed as invasive (Botryllus schlosseri and Bugula neritina) and 4 listed as a concern (Didemnum perlucidum, Lepas anserifera, Megabalanus rosa, and Tricellaria occidentalis). Their entire known range was categorised as introduced. Based on these established distributions, nearly 80 (XX%) records represented a range expansion of an NIS into a novel marine region (Table 1). Fewer than 15% of listed NIS were found within their native range (n = 104). This pattern only broke for the North Atlantic and Indian Oceans (Figure 3), though these regions were also four of the top five most common native ranges for the organisms found on AMD (Table 2).

For NIS found outside of their native range, over 80% were found in a marine region directly adjacent to their native range (n = 66). Connections between the nearest native marine region and the region in which  a debris item carrying an NIS was found showed a similar pattern in a graph analysis. Regions grouped using the leading eigenvector method [@citekey] formed three groups of stronger species transfer (Figure 4), in which the groups were majority composed of adjacent regions. A notable exception to this was in Group A, where the NW Pacific Ocean showed stronger connections with the North Atlantic Ocean and the Eastern Indian Ocean. Again, these exceptions were the top five most common native regions by species found on AMD (Table 2). 

Within regions, there appeared to be a statistically significant increase in the proportion of NIS in studies and on debris items found within 150km of cities of over 50,000 population (include stat and DF, p < 0.001) in a chi-squared test [@citekey]. The documented AMD community near urban regions was comprised of nearly 15% species listed in WRiMS, compared to just over 10% in communities on AMD found in areas more than 150km from urban areas (Table 3).

Summarised across locations and materials, the percent of NIS among the community of organisms on floating synthetic materials (12.3%) was shown to be sharply elevated above that in the comparable background marine community (0.2%) across the nine most common animal phyla. Communities randomly assembled from all taxonomically registered species in these nine phyla, representing the background marine community, showed significantly different (add test statistic, p < 0.001) proportions of NIS compared to communities randomly assembled from only those species found on AMD according to a logit-link ANOVA [@citekey]. Across 10,000 simulations, the randomly generated community of AMD showed a mean proportion of NIS equal to 12.3% (sd = 5.0%), compared to 0.3% (sd = 1.0%) background (Figure 5). A background marine community that was statistically similar to that found on AMD could be generated using weights between 39 and 81, according to the confidence intervals on a logit-link linear model (Figure 5). An identical proportion of NIS was generated when they were 64 times more likely to be selected. 


## Discussion


This analysis demonstrates that there is a statistically significant relationship between artificial substrates floating in the world’s oceans and organisms listed as NIS—a relationship commonly hypothesised by field researchers, but until now untested. From the first records of listed taxa on anthropogenic debris to appear in the scientific literature in 1994, this relationship has also remained relatively stable (Figure 1). Starting in 1997 with the first published identifications of NIS on debris, and throughout the topical research boom beginning in the early 2010’s, species listed as invasive, managed, or concerning in WRiMS have represented about 5% of the total number of unique species recorded on debris. This nearly 30-year stability in the overall ratio of NIS by species richness, which was also shown to be significantly higher than in the background marine community at a more granular level. Both may arise as a result of similarly steady driving forces such as methodological and reporting biases and biological influences.

Within the available data it was not possible to determine whether these forces were biological or methodological. Biologically, this relationship may arise from selection pressures inherent to synthetic materials floating at the surface of the ocean.  Anthropogenic marine debris may differ from common types of natural debris in ways that make it more suitable to the types of organisms simultaneously more likely to be listed as NIS.  For example, plastic debris provides comparatively negligible nutritional value to natural debris, while having a much higher surface energy, which may make settlement and maintenance of surface connections more difficult  (Andrady, 2015). Both field and lab studies have shown that differences such as these manifest as significantly different communities on natural and artificial marine debris (Dussud et al., 2018; Muthukrishnan, Al Khaburi, & Abed, 2018; Oberbeckmann, Kreikemeyer, & Labrenz, 2018).

The introduction and establishment patterns of the listed invasive species found on AMD further suggests the influence of biology on their elevated proportion within this community, and perhaps a relationship with marine urban areas. Most of the NIS found on AMD are known to foul the hulls of ships, marine infrastructure, and fishing and aquaculture equipment. Some are so successful at this that their native range is difficult to delineate due to the long history and persistence of their association, in particular, with international shipping, such as XX and XX. Other notable associations with artificial surfaces, such as coastal infrastructure and aquaculture equipment, are shown in the cases of XX, XX, XX, and XX. 

However, this observation could also arise from methodological factors. Shipping, infrastructure, and aquaculture are closely tied with human activities and economies, therefore species associated with them may be more commonly noted in association with economic impacts and harm to the environments humans frequent [@citekey]. This awareness may lead to more frequent listing as NIS. Listing as an NIS and associated prevention activities, including education and the availability of identification resources, may also result in increased familiarity and ability to identify these species by non-specialists. 

It was common within the OAAMD for the community to be reported as a value add within tangential studies, particularly studies related to measuring the volume or mass of debris itself. In these cases, it is unlikely that the entire AMD associated community was rigorously catalogued. Increased familiarity with NIS, and external incentives to identify and publish about them, may have increased the frequency at which NIS were reported relative to other taxa in these studies. Misidentifications as NIS may also have occurred in regions with morphologically similar native species, such as XX and XX [@citekey] due to non-specialist taxonomic identification methods.

Consideration of these methodological factors adds important context to the increased proportion of NIS found on anthropogenic debris. Within this context, though, there remains a total XX observations of XX unique listed invasive species on AMD. As an introduction vector, AMD may well be facilitating the spread of these listed invasive species. 

The significant relationship between urban areas and NIS on debris items suggests the ability of anthropogenic debris to enable secondary introductions to remote regions. Secondary introductions, or the transport and establishment of new populations of NIS outside of their initial location, significantly increase the difficulty of eliminating exotic species [@citekey]. This is particularly true when secondary introductions occur outside of known high-risk areas with frequent monitoring, like ports [@citekey]. The potential of AMD to facilitate passive secondary introductions may require NIS response plans for species known to foul the substrate to broaden their monitoring programs to include nearby regional beaches.  

Longer trips, such as those seen by Carlton et. al when studying debris items from the Japanese tsunami, might be far more likely during similar extreme events when large debris flows are quickly washed offshore. No other study has looked at this effect, although annual events such as monsoon and hurricane seasons may force debris out to sea in similar quantities.  Japanese tsunami debris overall showed XX% NIS among its community, a total of XX taxa, and XXX records of Mytilus galloprovincialis [@citekey], the Mediterranean mussel, across XX objects. Similar extreme events could also cause repeated introductions of genetically diverse and sexually mature organisms in a short time span, thereby sharply increasing the likelihood of successful establishment of an exotic species in a new region [@citekey]. 

In either scenario, AMD shows its potential as a vector for marine organisms to expand outside of their native range, and, potentially, cause ecological and economic harm to coastlines. It creates an abundance of habitat suitable for NIS transported through other, more highly regulated, vectors like shipping and aquaculture. Secondary introductions outside of monitored areas via the AMD vector may require a shift in marine biosecurity protocols and long-distance dispersal aboard AMD presents an entirely novel type of biosecurity threat. This study has established that not only are plastics and other synthetic materials capable of providing habitat to a diverse range of listed pest species, but that these species may be far more likely to be found in the AMD ecosystem.
